### 2018-05-09 thesis_report/Thesis_Background.tex
Background v1

The following chapter will go over prior work that closely relates to the topic of our research and areas of contribution.
We approach the definition of malware and how it can hard to detect and label by providing some background on the subject.
We then proceed to elucidate on different malware detection techniques, from simple methods to more complex approaches, focusing on a specific malware analysis technique that provided the basis for our work.

We advance to the general topic of \gls{ml}, after which we describe related work that studies the use of \gls{ml} in malware detection.
Lastly, we reference prior work that not only tackles our same goal, but also takes into account the impact of laboratory \textit{vs.} real world scenarios, as well as the notion of temporal based detection, or concept drift.

\newpage
\todo[inline]{CSF Related work}

We use this section to present prior work that closely relates to the topic of our research and our areas of contribution. Specifically, detecting malicious software by training supervised models on static information, and validation methodologies that resemble real-world conditions.

Shabtai et al.\cite{shabtai:survey} provide a survey directed
at the application of \gls{ml} classifiers to detect malware from static
features. Their work concerns the design and evaluation of such systems. Our contributions are inspired by the concern of how to correctly evaluate \gls{ml} classifiers, regarding size, reliability of labeling metrics and chronological evaluation.

In the topic of methodologies that resemble real-world conditions, Srndi\'c et al.\cite{vsrndic2013detection}
train and validate their malicious PDF detector under laboratory and real-world conditions. Laboratory conditions consist on applying regular cross-validation, whereas real-world conditions validate a newer dataset with a model created from outdated data (i.e.\ older then the validation), and also validate the model when the validation set spans one week and the training is gathered in the previous 4 weeks. They show that laboratory conditions inflate the results, when compared to real-world conditions.

Miller et al.\cite{miller:rev_int} also introduce sample temporal consistency. They show the impact of performance measurement technique on a dataset containing 1.1 million samples, when using cross-validation and temporally consistent samples. As noted by others, regular cross-validation showed inflated results when compared to temporally consistent samples.

Kolter et al.\cite{kolter:learning} learn to detect malicious executables in a dataset with under 4,000 samples, obtained from reliable sources, and evaluate their model under standard cross-validation and by gathering newer malware samples to validate for new and unseen samples. They show how the model provides optimal results under cross-validation, but for the unseen samples lower scores are obtained. Our work considers these results to compare how reliability affects performance.

Sebastián, M. et al.\cite{sebastian2016avclass} develop \textit{AVClass}, a tool that given a set of antivirus vendors, outputs the most likely family name. They test their tool under 10 datasets, totaling 8.9 M samples, with results showing an F1 measure up to 93.9 on labeled datasets. The tool takes as input the labels as seen in VirusTotal, tokenizes the labels, replaces known aliases and general names (e.g.\ win32, trojan, generic), ending with possible family names. These remaining names are counted and the most frequent one is given as family name. Our work takes advantage of this tool, not to label malware families, but minimal modifications, to label more general malware classes (e.g.\ trojan, virus).

Deo, A. et al.~\cite{deo2016prescience} focus over the problem of \gls{ml} models becoming antiquated over time, given how malware evolves. This problem is seen as concept drift, where the performance of a model over time diminishes, as the statistical properties of malware (i.e.\ features) change over time. They study how probabilistic predictors can help minimize the aforementioned problem, by indicating when retraining of a given model is necessary.

Both ~\cite{deo2016prescience} and ~\cite{jordaney2017transcend} are on the subject of our work, regarding the problem of concept drift, but differ on the study-case. In our work we acknowledge the concept but focus on its relative effects and how one can balance the size of needed training data vs. validation data, when maintaining a temporal consistent dataset. Whereas their work provide indicators for when should a model be retrained when the problem of concept drift becomes significant.

\newpage
\todo[inline]{Project Report related work}

@@

Malware, which can be referred to by other names like malicious software or malicious code, can be described as \say{a program that has malicious intent}\cite{christodorescu:semantics}, or more detailed as any code added, changed, or removed from a software system in order to intentionally cause harm or subvert the intended function of the system\cite{mcgraw:mal_code}.

The term malware was first used by Yisrael Radai in 1990\cite{elisan:malware}, before then malicious software was referred to as computer viruses, a notion which was first formalized by Cohen in 1983\cite{cohen:virus}. Given term computer virus predates the term malware, it is not uncommon to see both terms used interchangeably. For the purpose of this work the term malware is adopted, for reasons explained in the following paragraphs.

To the exponential growth of malware\cite{av-test:report}, the ability to distinguish between malware samples is crucial. In the biological world, different types of infections reckon distinctive disinfections, the same applies to the digital world. The infection method, behavior and subsequently purpose of malware samples vary, hence being able to make a distinction facilitates prevention and disinfection by anti-virus solutions.

@@

With respect to the propagation method, malware can be split into three classes\cite{kolter:learning}.

\paragraph{Virus.}\footnote{This work uses virus as a subset of malicious code, hence the choice of using the term malware over virus.}This type of malware inherits its name from resembling to biological viruses. A virus is usually composed by two main subroutines. The first subroutine is responsible for infecting other programs by attaching the virus code, while the second subroutine is the actual malware payload that contains the malware purpose (i.e. virus payload)\cite{chen:evolution}. Viruses are attached to programs and propagate when an infected program is run, hence depending on other programs (as hosts) and user interaction. For viruses to infect new systems, an infected file must be carried between them by a user (e.g. USB pen, email attachment), reinforcing the user's role in this propagation method.

\paragraph{Worm.} The increasing number of network connected devices facilitates the application of worms to propagate malware. A worm shares the self-replicating ability of viruses, but discards user interaction and the need for a host. This is possible by making a worm an independent program that exploits networks to find vulnerable systems to infect with a copy of themselves\cite{chen:evolution}. Similarly to viruses, worms can contain additional payload to perform malicious actions on infected systems (e.g. Blaster worm exploited a vulnerability in RPC for propagation while its payload flooded a Microsoft domain\footnote{W32/Blaster worm, CERT, August 11, 2003 [http://www.cert.org/historical/advisories/CA-2003-20.cfm]}).

\paragraph{Trojan.} While viruses and worms focus on self-propagation, trojans focus on deceiving users into executing them, disregarding propagation. Trojans try to appeal users with some useful functionality as to allure into running the program\cite{szor:art}. By hoaxing users into running them, trojans bypass the need for a host (in the virus case) or an exploit (in the worm case) to perform malicious actions.\\

These three classes are not exclusive and can be used in conjunction to facilitate propagation (e.g. a virus that relies on a trojan to start propagating). Given the inexistent standards for malware naming, some literature\cite{szor:art} define worms as a subset of viruses, as some worms start by attaching to files. With the general infection methods layered out, the following will succinctly describe the general naming when referring to the malware purpose (i.e. its payload).

Given the high numbers of malware multiple purposes imply multiple names, as such the following lists some of the most commonly heard names and definitions as given by McAfee\cite{mcafee:glossary}:

\begin{itemize}
	\item \textit{Adware}: Program that automatically displays advertisement to the user. Most do not cause direct harm.
	\item \textit{Backdoor}: Installs or takes advantage of an unknown entry point to a user's system, allowing remote control of a system.
	\item \textit{Dropper}: Program that facilitates the introduction of other malware instances into a system.
	\item \textit{Keylogger}: Program that monitors a user's input (e.g. keyboard strokes) to steal private information.
	\item \textit{Password Stealer (PWS)}: Program that specifically targets a user's personal information, like usernames and passwords.
	\item \textit{Ransomware}: Program that restricts access to a system/files (usually by encryption) and demands a ransom to allow a user his access.
	\item \textit{Rootkit}: Program designed to hide the existence of other (malicious) programs, hindering detection of malicious applications.
	\item \textit{Scareware}: Program that fakes the existence of malware and tricks the user into paying for disinfection or into downloading real malware.
	\item \textit{Spyware}: Program that spies on a user's activity through multiple vectors (e.g. taking prints of a user's desktop/webcam).
\end{itemize}

@@

One of the first attempts was made in 1991, by the \gls{caro}\cite{caro:naming}. This attempt was aimed at viruses only and focused on discriminating and relating virus samples. The convention defines an hierarchical naming with three levels. The base of the hierarchy is the family name, where structurally similar virus should be grouped. To further distinguish instances, a group name and major variant name is applied. To differentiate versions of the same instance and provide additional information about a sample, minor variant name and modifier fields are to be used. In sum, a virus name should look as following:

@@

To enable differentiation from other types of malware, an improvement to \gls{caro}'s naming was proposed in 1999 by Geral Scheidl\cite{scheidl:naming}. Scheidl's proposal included a prefix that consisted of the platform targeted by the malware and a type (i.e. previously defined classes and names), rendering a more in-depth name:

@@

In the Virus Bulletin magazine of January 2002\cite{virus_bulletin} the naming problem is again addressed, referencing that only a small number of vendors apply \gls{caro}'s names and that the lack of cooperation between anti-virus vendors lead to multiple aliases for malware names. Again a naming convention is proposed with focus on detail by using three levels of information: malware name, classification and full text description.

@@

In section \ref{sec:introduction} a brief description of malware detection methods was given. This section will provide a deeper description of detection methods, from the simplest techniques to more complex ones. The following is based on the work by Peter Szor, \textit{The Art of Computer Virus Research and Defense}\cite{szor:art}.

@@

\subsubsection{First-Generation Scanners.} These type of scanners work at a fairly simple level and focus searching for known byte sequences (i.e. known to be malware sequences). The applied principles include:

@@

\paragraph{Mismatches.} Also similar to string scanning, yet there can be $N$ number of bytes in the string that can take any value and in any position (e.g. the string \say{\texttt{01 02}} with mistmatch of 1 matches \say{\texttt{01 AA}} and \say{\texttt{AA 02}}).

\paragraph{Generic Detection.} This method typically applies both wildcards and mismatches to create generic strings that can scan for several variants of a family of malware.

\paragraph{Hashing.} This term is not a method by itself, but techniques that speed up searching algorithms, enhancing previous methods.

\paragraph{Bookmarks.} This method is both used to provide more accurate detections and disinfections. Bookmarks provide specific locations for where strings are to be searched (i.e. offsets).

@@

\paragraph{Hyperfast Disk Access.} This last technique also provides a speed improvement. It does so by bypassing operating system level \gls{api} to read the disk directly with the \gls{bios}. It is worth noting that this technique is hard to implement nowadays due to the high variety of file systems and disk controllers.\\

The described methods, excluding the ones that focus on speed improvement, can be assigned to the signature-based type of methods, as they depend on searching some previously defined byte sequence to be effective. Their effectiveness depends on an updated knowledge base with the strings to be matched.

\subsubsection{Second-Generation Scanners.} These methods improve on the previous generation by doing an analysis at a higher level of abstraction. Where first-generation scanners would match sequences of bytes, second-generation scanners also look at the bytes representation. The applied principles include:

\paragraph{Smart Scanning.} This method takes the assembly representation of the bytes into account and instructions that have no effect (e.g. \texttt{NOP}) can be skipped. The resulting signature is smaller and more easily detects mutations. The same principle can be applied to textual malware (e.g. scripts), where extra white spaces have no effect.

\paragraph{Skeleton Detection.} This method is specifically useful in detecting textual malware (e.g. macros) families. By parsing the macro statements line by line, the scanner can drop nonessential statements. The result can be seen as a skeleton of the full macro that contains only the relevant code that appears in malware.

\paragraph{Nearly Exact Identification.} This method provides a more accurate way to detect malware. First-generation scanners base detection mainly on a byte sequence, whereas nearly exact identification makes use of double-string detection. This is specifically useful to identify same family malware. Another nearly exact identification method applies a checksum to a range from the malware body, the resulting checksum can then be used to detect and disinfection files.

\paragraph{Exact Identification.} This method enhances the previous method by adding the ability to differentiate malware variants. It does so by using as many checksum ranges as needed to identify variants of the same malware, improving the accuracy and facilitating disinfection.\\

@@

\subsubsection{Algorithmic Scanning Methods.} These methods are not so focused on direct matching of signatures, but on the routines implemented in the scanner that allow the detection of malware. Early implementations of algorithmic scanning were hard-coded in the antivirus, but with different types and families emerging, the hard-coded algorithms could easily become obsolete. To overcome this, vendors would introduce more specific routines for malicious programs when updating the antivirus. Examples of these methods are filtering, where signatures are only applied if an object meets a particular filter; static decryptor, which detects encrypted malware; and X-ray, which tries to decrypt malware that use simple encryption methods (e.g. \say{\texttt{XOR}} with constant key).

@@

\subsubsection{Code Emulation.} This detection technique uses dynamic analysis (i.e. analysis during runtime) to detect malware by simulating code execution. A virtual machine is used to simulate the system's components, thus the malicious files are encapsulated and no code is executed by the real processor. Code emulation can be faster than using algorithmic scanning methods when dealing with encrypted/obfuscated malware, as the code can be emulated and the decrypted/deobfuscated code is much easier to obtain. The downside to emulation comes from a time perspective, as more resources are needed to emulate code.

\subsubsection{Heuristic Analysis.} This detection technique takes into account several characteristics of malicious software, both static and dynamic. Examples of such characteristics are code execution starting in uncommon sections, suspicious code redirection or incorrect size of code in header. Using the previously described methods, heuristic analysis can take into account different characteristics that are present or absent to assert if a file is deemed malicious or not. The upside of this technique is that it allows for never seen malware to be correctly flagged, but brings the downside of a higher amount of false positives.

\subsubsection{Sand-Boxing.} This detection technique builds on the already mentioned code emulation, but with a higher level of complexity. With sand-boxing the whole system is emulated (i.e. virtualized) and suspicious or malicious programs have access to a copy of the real resources, effectively limiting the effects of the program and consequently protecting the system.

This technique allows to extract dynamic information without compromising the real system, which can then be both used to detect if a suspicious program is malware, by applying the aforementioned techniques, and to help analyze new types of malware. This technique demands more resources then previous ones and may not be ideal to use inside an anti-virus application. Other downsides of sand-boxing come as compatibility issues, the possibility of malware detecting virtualization, which can hinder analysis, or even if the virtualization software is vulnerable and malicious code can escape the sandbox into the host machine.\\

Detecting malware is not as simple as applying one of the mentioned techniques, different methods work best in different scenarios, hence antivirus need to balance the usage of these techniques to keep systems protected.

@@

Building on the last described method of sand-boxing, this section will describe a specific implementation of sand-boxing, namely Cuckoo. Cuckoo Sandbox is a malware analysis system created by Claudio Guarnieri\cite{tool:cuckoo}.

When analyzing a malware sample, an analyst will have to create his own virtual environment, transfer the sample into that environment and manually interact with the sample to see how it behaves and what it does. Cuckoo facilitates analysis by automating this process. The Cuckoo system will take the sample to analyze and run it inside a predefined virtual environment (e.g. Windows XP). The system is then capable of simulating user interaction, take screenshots of the environment, extract static and dynamic information about a sample, monitor network traffic and perform memory analysis. The time spent doing the analysis can be configured by the user. Upon finishing the analysis the system provides reports about the sample. Examples of what is contained in these reports include: what functions were imported from which \gls{dll}, what processes were created and the \gls{api} calls order and which files were created/modified.

The creator of Cuckoo and one of its developers, Alessandro Tanasi, also created a free malware analysis service, Malwr\cite{tool:malwr}. This service implements the Cuckoo system and allows users to upload their own files, which are then run with Cuckoo. The service is presented as a web application that takes in samples and returns a report with Cuckoo's results. An example of the behavioral analysis (i.e. created process and \gls{api} calls) can be seen in figure \ref{fig:malwr_sample}.

%\begin{figure}[h]
%	\includegraphics[width=\textwidth]{malwr_sample}
%	\caption{Example of a behavioral report from Malwr}
%	\label{fig:malwr_sample}
%\end{figure}

For the purpose of detecting malware via \gls{ml}, the reports provided in Malwr supply various information that can be used to fulfill the current work. The whole website content is provided under \gls{cc} CC-BY-NC-SA 3.0\cite{cc-by-nc-sa} license, allowing the use of the data for this work's purpose, as long as credits are maintained.

@@

Even before the modern concept of computer proposed by Alan Turing, mankind has always searched for ways to either automate or facilitate more complex tasks. With the technological advances that provide higher computation power, computer science fields like \gls{ai} have greatly improved in recent years. One of the \gls{ai} fields that has benefit from the technological advances is \gls{ml}. The current subsection will provide a more detailed description of \gls{ml} and its use for the current work.

\gls{ml} is a subset of \gls{ai} in which algorithms are applied to learn and make predictions on data, or more formally defined by Tom Mitchell\cite{mitchell:ml}:\\
\say{\emph{A computer program is said to learn from experience E with respect
		to some class of tasks T and performance measure P, if its performance at tasks in
		T, as measured by P, improves with experience E.}}

A noteworthy nuance about the given definition, and the \gls{ml} field in general, is that algorithms are not designed to deal with specific problems. Meaning that as long as a learning problem is well defined through $T$, $P$ and $E$, an algorithm based on these three features is able to learn, but not necessarily solve a problem.

After defining a learning problem, to obtain a useful outcome (i.e. model) one or more learning algorithms are applied. Learning algorithms fall within one of the following classes: supervised learning; unsupervised learning; semi-supervised learning or reinforcement learning. Each of the categories is defined on how the program learns the problem. A more detailed description, as seen in \cite{norvig:ai} is as follows:

\subsubsection{Supervised Learning.} In this category the program is given a corpus (i.e. experience) with input-output pairs and it must learn how to map a never seen input into the correct output. Within supervised learning, a learning problem can be further classified into a classification problem, if the output is a set of finite values (e.g. is it going to rain tomorrow?), or into a regression problem, if the output is continuous (e.g. how likely is it to rain tomorrow?).

\subsubsection{Unsupervised Learning.} In this category the program is given a corpus and it must learn patterns from the input, hence no correct or wrong output is expected, only how different input relates. The most common task in unsupervised learning is clustering, where input is grouped based on similar characteristics.

\subsubsection{Semi-supervised Learning.} This category is a combination of the previous ones and exists because sometimes the learning problem is based on a dataset that combines both labeled and unlabeled inputs or on a dataset that contains noisy data. The combination of supervised and unsupervised learning provides a better overall result in such scenarios.

>>>>

The following chapter will go over prior work that closely relates to the topic of our research and areas of contribution.
We approach the definition of malware and how it can be hard to detect and label by providing some background on the subject.
We then proceed to elucidate on different malware detection techniques, from simple methods to more complex approaches, focusing on a specific malware analysis technique that provided the basis for our work.

We advance to the general topic of \gls{ml}, after which we describe related work that studies the use of \gls{ml} in malware detection.
Lastly, we reference prior work that not only tackles our same goal, but also takes into account the impact of \textit{laboratory} \textit{vs.} \textit{real world} scenarios, as well as the notion of temporal based detection, or concept drift.

@@

The term malware was first used by Yisrael Radai in 1990~\cite{elisan:malware}, before then malicious software was referred to as computer viruses, a notion which was first formalized by Cohen in 1983~\cite{cohen:virus}.
Given term computer virus predates the term malware, it is not uncommon to see both terms used interchangeably.
% For the purpose of this work the term malware is adopted, for reasons explained in the following paragraphs.

To the exponential growth of malware~\cite{av-test:report}, the ability to distinguish between malware samples is crucial.
In the biological world, different types of infections reckon distinctive disinfections, the same applies to the digital world.
The infection method, behavior and subsequently purpose of malware samples vary, hence being able to make a distinction facilitates prevention and disinfection by anti-virus solutions.

@@

With respect to the propagation method, malware can be split into three classes~\cite{kolter:learning}:

\paragraph{Virus.}%\footnote{This work uses virus as a subset of malicious code, hence the choice of using the term malware over virus.}
This type of malware inherits its name from resembling to biological viruses.
A virus is usually composed by two main subroutines.
The first subroutine is responsible for infecting other programs by attaching the virus code, while the second subroutine is the actual malware payload that contains the malware purpose (\ie\ virus payload)~\cite{chen:evolution}.
Viruses are attached to programs and propagate when an infected program is run, either consciously (\eg\ clicking on executables) or unconsciously (\eg\ auto-run features), hence depending on other programs (as hosts) and user interaction.
For viruses to infect new systems, an infected file must be carried between them by a user (\eg\ USB pen, email attachment), reinforcing the user's role in this propagation method.

\paragraph{Worm.} The increasing number of network connected devices facilitates the application of worms to propagate malware.
A worm shares the self-replicating ability of viruses, but discards user interaction and the need for a host.
This is possible by making a worm an independent program that exploits networks to find vulnerable systems to infect with a copy of themselves~\cite{chen:evolution}.
Similarly to viruses, worms can contain additional payload to perform malicious actions on infected systems (\eg\ Blaster worm exploited a vulnerability in RPC for propagation while its payload flooded a Microsoft domain\footnote{W32/Blaster worm, CERT, August 11, 2003 [http://www.cert.org/historical/advisories/CA-2003-20.cfm]}).

\paragraph{Trojan.} While viruses and worms focus on self-propagation, trojans focus on deceiving users into executing them, disregarding propagation.
Trojans try to appeal users with some useful functionality as to allure into running the program~\cite{szor:art}.
By hoaxing users into running them, trojans bypass the need for a host (in the virus case) or an exploit (in the worm case) to perform malicious actions.

\medskip

These three classes are not exclusive and can be used in conjunction to facilitate propagation (\eg\ a virus that relies on a trojan to start propagating).
Given the inexistent standards for malware naming, some literature~\cite{szor:art} define worms as a subset of viruses, as some worms start by attaching to files.
With the general infection methods layered out, the following will succinctly describe the general naming when referring to the malware purpose (\ie\ its payload).

Given the high numbers of malware multiple purposes imply multiple names, as such the following lists some of the most commonly heard names and definitions as given by McAfee~\cite{mcafee:glossary}:

\begin{itemize}
	\item \textit{Adware}: Program that automatically displays advertisement to the user. Most do not cause direct harm.
	\item \textit{Backdoor}: Installs or takes advantage of an unknown entry point to a user's system, allowing remote control of a system.
	\item \textit{Bot}: Short for \say{robot}; A program that receives commands from a cybercriminal and executes them blindly, due to this characteristic, compromised computers are called zombies.
	\item \textit{Dropper}: Program that facilitates the introduction of other malware instances into a system.
	\item \textit{Keylogger}: Program that monitors a user's input (\eg\ keyboard strokes) to steal private information.
	\item \textit{Password Stealer (PWS)}: Program that specifically targets a user's personal information, like usernames and passwords.
	\item \textit{Ransomware}: Program that restricts access to a system/files (usually by encryption) and demands a ransom to allow a user his access.
	\item \textit{Remote Administration Tool - RAT}: Program designed to give an administrator remote control of a system. Similar to \textit{Backdoor} and \textit{Bot}, but usually has higher privileges on infected systems.
	\item \textit{Rootkit}: Program designed to hide the existence of other (malicious) programs, hindering detection of malicious applications.
	\item \textit{Scareware}: Program that fakes the existence of malware and tricks the user into paying for disinfection or into downloading real malware.
	\item \textit{Spyware}: Program that spies on a user's activity through multiple vectors (\eg\ taking prints of a user's desktop/webcam).
\end{itemize}

@@

One of the first attempts was made in 1991, by the \gls{caro}~\cite{caro:naming}.
This attempt was aimed at viruses only and focused on discriminating and relating virus samples.
The convention defines an hierarchical naming with three levels.
The base of the hierarchy is the family name, where structurally similar virus should be grouped.
To further distinguish instances, a group name and major variant name is applied.
To differentiate versions of the same instance and provide additional information about a sample, minor variant name and modifier fields are to be used.
In sum, a virus name should look as following:

@@

To enable differentiation from other types of malware, an improvement to \gls{caro}'s naming was proposed in 1999 by Geral Scheidl~\cite{scheidl:naming}.
Scheidl's proposal included a prefix that consisted of the platform targeted by the malware and a type (\ie\ previously defined classes and names), rendering a more in-depth name:

@@

In the Virus Bulletin magazine of January 2002~\cite{virus_bulletin} the naming problem is again addressed, referencing that only a small number of vendors apply \gls{caro}'s names and that the lack of cooperation between anti-virus vendors lead to multiple aliases for malware names.
Again a naming convention is proposed with focus on detail by using three levels of information: malware name, classification and full text description.

@@

This section will provide a deeper description of detection methods, from the simplest techniques to more complex ones. The following is based on the work by Peter Szor, \textit{The Art of Computer Virus Research and Defense}~\cite{szor:art}, which although not recent, provides a solid background on detection methods.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{First-Generation Scanners}
\label{subsection:first_gen_scanners}

These type of scanners work at a fairly simple level and focus searching for known byte sequences (\ie\ known to be malware sequences).
The applied principles include:

@@

\paragraph{Mismatches.} Also similar to string scanning, yet there can be $N$ number of bytes in the string that can take any value and in any position (\eg\ the string \say{\texttt{01 02}} with mistmatch of 1 matches \say{\texttt{01 AA}} and \say{\texttt{AA 02}}).

\paragraph{Generic Detection.} This method typically applies both wildcards and mismatches to create generic strings that can scan for several variants of a family of malware, hence being more robust to subtle changes in malware.

\medskip

These principles can be assigned to the signature-based type of methods, as they depend on searching some previously defined byte sequence to be effective.
Their effectiveness depends on an updated knowledge base with the strings to be matched.

Alongside these basic principles, the following can be used in conjunction to improve the scanning speed and reliability:

\paragraph{Hashing.} This term is not a method by itself, but techniques that speed up searching algorithms, enhancing previous methods.
It takes advantage of the hash table data structure to match malware signatures.

\paragraph{Bookmarks.} This method is both used to provide more accurate detections and disinfections. Bookmarks provide specific locations for where strings are to be searched (\ie\ offsets).

@@

\paragraph{Hyperfast Disk Access.} This last technique also provides a speed improvement.
It does so by bypassing operating system level \gls{api} to read the disk directly with the \gls{bios}.
In addition to improve the scanners performance, this principle is also useful as an anti-stealth technique, given it works below the filesystem layer.
It is worth noting that this technique is hard to implement nowadays due to the high variety of file systems and disk controllers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Second-Generation Scanners}
\label{subsection:second_gen_scanners}

These methods improve on the previous generation by doing an analysis at a higher level of abstraction.
Where first-generation scanners would match sequences of bytes, second-generation scanners also look at the bytes representation.
The applied principles include:

\paragraph{Smart Scanning.} This method takes the assembly representation of the bytes into account and instructions that have no effect (\eg\ \texttt{NOP}) can be skipped.
The resulting signature is smaller and more easily detects mutations.
The same principle can be applied to textual malware (\eg\ scripts), where extra white spaces have no effect.

\paragraph{Skeleton Detection.} This method is specifically useful in detecting textual malware (\eg\ macros) families.
By parsing the macro statements line by line, the scanner can drop nonessential statements.
The result can be seen as a skeleton of the full macro that contains only the relevant code that appears in malware.

\paragraph{Nearly Exact Identification.} This method provides a more accurate way to detect malware.
First-generation scanners base detection mainly on a byte sequence, whereas nearly exact identification makes use of double-string detection.
The principle comes from using two strings to match a given malware, if only one of those strings match, then it is likely to be a variant of a known malware.
This is specifically useful to identify same family malware.
Another nearly exact identification method applies a checksum to a range from the malware body, this has the added benefit of better accuracy without overloading the anti-virus database.

\paragraph{Exact Identification.} This method enhances the previous method by adding the ability to differentiate malware variants.
It does so by using as many checksum ranges as needed to identify variants of the same malware, improving the accuracy and facilitating disinfection.

\medskip

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Algorithmic Scanning Methods}
\label{subsection:algorithmic_scanning}

These methods are not so focused on direct matching of signatures, but on the routines implemented in the scanner that allow the detection of malware.
Early implementations of algorithmic scanning were hard-coded in the anti-virus, but with different types and families emerging, the hard-coded algorithms could easily become obsolete.
To overcome this, vendors would introduce more specific routines for malicious programs when updating the anti-virus.

Examples of these methods are filtering, where signatures are only applied if an object meets a particular filter; static decryptor, which detects encrypted malware; and X-ray, which tries to decrypt malware that use simple encryption methods (e.g. \say{\texttt{XOR}} with constant key).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Code Emulation}
\label{subsection:code_emulation}

This detection technique uses dynamic analysis (\ie\ analysis during runtime) to detect malware by simulating code execution.
A virtual machine is used to simulate the system's components, thus the malicious files are encapsulated and no code is executed by the real processor.

Code emulation can be faster than using algorithmic scanning methods when dealing with encrypted/obfuscated malware, as the code can be emulated and the decrypted/deobfuscated code is much easier to obtain.
The downside to emulation comes from a time perspective, as more resources are needed to emulate code.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Heuristic Analysis}
\label{subsection:heuristic_analysis}

This detection technique takes into account several characteristics of malicious software, both static and dynamic.
Examples of such characteristics are code execution starting in uncommon sections, suspicious code redirection or incorrect size of code in header.

Using the previously described methods, heuristic analysis can take into account different characteristics that are present or absent to assert if a file is deemed malicious or not.

The upside of this technique is that it allows for never seen malware to be correctly flagged, but brings the downside of a higher amount of false positives.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Sand-Boxing}
\label{subsection:sand_boxing}

This detection technique builds on the already mentioned code emulation, but with a higher level of complexity.
With sand-boxing the whole system is emulated (\ie\ virtualized) and suspicious or malicious programs have access to a copy of the real resources, effectively limiting the effects of the program and consequently protecting the system.

This technique allows to extract dynamic information without compromising the real system, which can then be used to detect if a suspicious program is malware, by applying the aforementioned techniques, and to help analyze new types of malware.
This technique demands more resources then previous ones and may not be ideal to use inside an anti-virus application.

Other downsides of sand-boxing come as compatibility issues, the possibility of malware detecting virtualization, which can hinder analysis, or even if the virtualization software is vulnerable and malicious code can escape the sandbox into the host machine.

\medskip

Detecting malware is not as simple as applying one of the mentioned techniques, different methods work best in different scenarios, hence anti-virus need to balance the usage of these techniques to keep systems protected.

@@

Building on the last described method of sand-boxing, this section will describe a specific implementation of sand-boxing, namely Cuckoo.
Cuckoo Sandbox is a malware analysis system created by Claudio Guarnieri~\cite{tool:cuckoo}.

When analyzing a malware sample, an analyst will have to create his own virtual environment, transfer the sample into that environment and manually interact with the sample to see how it behaves and what it does.
Cuckoo facilitates analysis by automating this process.

The Cuckoo system will take the sample to analyze and run it inside a predefined virtual environment (\eg\ Windows XP).
The system is then capable of simulating user interaction, take screenshots of the environment, extract static and dynamic information about a sample, monitor network traffic and perform memory analysis.
The time spent doing the analysis can be configured by the user.
Upon finishing the analysis the system provides reports about the sample. The reports' information can be divided into 5 sections:
\begin{itemize}
	\item \textbf{Quick Overview} which includes information like: file type and details; analysis duration; signatures (\ie\ predefined patterns Cuckoo deems suspicious); screenshots during runtime; hosts and domains contacted; files, mutexes and registries used.
	\item \textbf{Static Analysis} which includes information like: file version information; file sections; resources (\eg\ icons); statically imported libraries and their functions; output of the \texttt{strings}\footnote{Linux binary that outputs the strings of printable characters in files.} command on the file; VirusTotal~\cite{tool:virustotal} classification report.
	\item \textbf{Behavioral Analysis} which includes information like: spawned processes; \gls{api} call order for each process.
	\item \textbf{Network Analysis} which includes information like: domains and their resolved IPs; HTTP requests; IRC traffic; SMTP traffic.
	\item \textbf{Dropped Files} which includes information about files created by the sample: their size, name and hash.
\end{itemize}

Some of the defined sections are not always available.
Their availability depends both on a successful analysis by Cuckoo, as well as the type of submitted file.
For example, only binary executables (\eg\ \gls{dll}, \gls{pe}) provide static analysis information.

The creator of Cuckoo and one of its developers, Alessandro Tanasi, also created a free malware analysis service, Malwr~\cite{tool:malwr}.
This service implements the Cuckoo system and allows users to upload their own files, which are then run with Cuckoo.
The service is presented as a web application that takes in samples and returns a report with Cuckoo's results.
An example of the behavioral analysis (\ie\ created process and \gls{api} calls) can be seen in Figure \ref{fig:malwr_sample}.

\begin{figure}[!htb]
	\includegraphics[width=\columnwidth]{Figures/malwr_sample.png}
	\caption{Example of a behavioral report from Malwr.}
	\label{fig:malwr_sample}
\end{figure}

Unfortunately at the time of writing, Malwr is under maintenance hence the provided image will be outdated once the new version is live.

For the purpose of detecting malware via \gls{ml}, the reports provided in Malwr supply various information that can be used for the current work.
The whole website content is provided under \gls{cc} CC-BY-NC-SA 3.0\cite{cc-by-nc-sa} license, allowing use of the data for this work's purpose, as long as credits are maintained.

@@

Even before the modern concept of computer proposed by Alan Turing, mankind has always searched for ways to either automate or facilitate more complex tasks.
With the technological advances that provide higher computation power, computer science fields like \gls{ai} have greatly improved in recent years.

One of the \gls{ai} fields that has benefit from the technological advances is \gls{ml}.
The current subsection will provide a more detailed description of \gls{ml} and its use for the current work.

\gls{ml} is a subset of \gls{ai} in which algorithms are applied to learn and make predictions on data, or more formally defined by Tom Mitchell~\cite{mitchell:ml}:
\begin{center}
\say{\emph{A computer program is said to learn from experience $E$ with respect to some class of tasks $T$ and performance measure $P$, if its performance at tasks in $T$, as measured by $P$, improves with experience $E$.}}
\end{center}

A noteworthy nuance about the given definition, and the \gls{ml} field in general, is that these algorithms are not designed to deal with specific problems.
Meaning that as long as a learning problem is well defined through $T$, $P$ and $E$, an algorithm based on these three features is able to learn, but not necessarily solve any problem.
This creates a contrast to \gls{ai} fields, where algorithms are built for specific tasks.

After defining a learning problem, to obtain a useful outcome (\ie\ model) one or more learning algorithms are applied.
Learning algorithms fall within one of the following classes: \textit{supervised learning}; \textit{unsupervised learning}; \textit{semi-supervised learning} or \textit{reinforcement learning}.

Each of the categories is defined on how the problem is learned . A more detailed description, as seen in \cite{norvig:ai} is as follows:

\paragraph{Supervised Learning.} In this category the program is given a corpus (\ie\ experience) with input-output pairs and it must learn how to map a never seen input into the correct output.
Within supervised learning, a learning problem can be further classified into a classification problem, if the output is a set of finite values (\eg\ is it going to rain tomorrow?), or into a regression problem, if the output is continuous (\eg\ how likely is it to rain tomorrow?).

\paragraph{Unsupervised Learning.} In this category the program is given a corpus and it must learn patterns from the input, hence no correct or wrong output is expected, only how different input relates.
The most common task in unsupervised learning is clustering, where input is grouped based on similar characteristics.

\paragraph{Semi-supervised Learning.} This category is a combination of the previous ones and exists because sometimes the learning problem is based on a dataset that combines both labeled and unlabeled inputs or on a dataset that contains noisy data. 
The combination of supervised and unsupervised learning provides a better overall result in such scenarios.
